---
title: "Velib_Notebook"
output: html_notebook
---

# Why getting data of citybikes ?
Public transportations have made a huge progress these last years especially when it comes to green mobility.Indeed, Public Leasing Bikes are flourishing in urban areas for offering a fast and environmental friendly way to travel for people.As the demand increases while public powers are putting efforts in developing bikelanes instead of roads. However if it is growing too fast , city centers could face local shortages if every city area is not well equipped with enough citybikes. 

This data analysis project is conducted over an hourly updated dataset of Paris citybikes called Velib. That is why in this exploratory analysis shortages per areas (18 totally in Paris) will be showcased for drawing some recommendations on further locations to build. 

## Dataset Description 
Paris has more than 1400 stations and more than 400km2 covered with mechanical and ebikes. Data is updated in real time what allows to see the evolution of service in terms of available bikes for each station. 

Here are listed below the variables of this dataset : 
1. stationCode : chr 
2. station_id : num
3.num_bikes_available: int  (Number of bikes available)
4. mechanical : int  (Number of mechanicald Bikes available)
5. ebike : int  (Number of Ebikes available)
6. numBikesAvailable : int  (Total Number of bikes available)
7. num_docks_available: int  (Number of docks available)
8. numDocksAvailable : int  
9. is_installed : int (Wether if the station is installed or not 1-yes 0-no)
10. is_returning : int  (1- bike is returned 0-bike is not)
11. is_renting : int (1-bike is rented 0-bike is not)
12. last_reported : int (timestamp when the situation was reported)
13. name : chr (name of the station)
14. lat : num (latitude of the station)
15. lon : num (longitude of the station)
16. capacity : num (total capacity of the station)
17. rental_methods : (NA refers to subscription CREDIT CARD to one time payment)

Quick explaination of the dataset : the above aggregated data is directly obtained by GET request from public APIs.Each time the API is called the sent response is capturing the situation of bike trafic in Paris. As bike come and go the number of bikes available and dock available varies depending on the number of bikes actually rented during two time periods. One line equals to one transaction when each time a bike is rented.

For this analysis we will only take in account the transactions made in year 2022.

```{r}
# connect to SQLite Database
conn <- dbConnect(RSQLite::SQLite(), "Velib_DB.db")

velib <- dbGetQuery(conn, 'SELECT *
                          FROM Velib_Data') %>%
  tibble()

# removing NA values
velib <- velib %>%
  na.omit()

#as station id and station code are numerics let's convert it in strings
velib$station_id  <- as.character(velib$station_id)
velib$stationCode <- as.character(velib$station_id)

#as the same date has to be converted from an int to a date format
velib$last_reported <- as.POSIXct(velib$last_reported, origin="1970-01-01")

head(velib,5)

```
As you might have noticed the oldest transaction reported were in 2018 ? Obviously we can formulate some hypothesis about this. Every user is not peculiar when it comes to ride or park a bike. And some have seen their beloved ones ending in the waters of the Seine or broken by accident.

Life is though for citybikes :), but numerous ones are still remaining and are used everyday without being seriously damaged. That is why remaining operational bikes will only be taken in account in this study. 

```{r}
#filter by year 2022
velib <- velib %>%
  filter(as.Date(velib$last_reported) >= "2022-01-01")
```

### Overview

```{r}
station <- dbGetQuery(conn, 'SELECT COUNT(DISTINCT station_id) as count_station
                               FROM Velib_Data')

total_bikes <- dbGetQuery(conn,'SELECT DISTINCT station_id,
                      SUM(num_bikes_available) as total_count,
                      SUM(mechanical) as count_mechanical,
                      SUM(ebike) as ebike_count
                      FROM Velib_Data') %>%
                      .[,-1]
station
total_bikes
```
The 1452 stations are counting 423338 bikes with 265304 mechanical and 158084 ebikes over the fleet. Mechanical bikes are representing more than 2/3 of the bikes population in Paris !

### What is the most common rental method ?
Citybikes can be easily used with city transportation card with a significant discount on all subscription. So we might think that bikes are more rented by susbcription rather than with Credit Card which means that the user is unlocking the bike for a single ride. 

```{r}

payment <- dbGetQuery(conn, 'SELECT rental_methods, count(*) as subscription
                             FROM Velib_Data
                             WHERE rental_methods == "NA"
                             UNION
                             SELECT rental_methods, count(*) as subscription
                             FROM Velib_Data
                             WHERE rental_methods == "CREDITCARD"')
  
payment
```
```{r}
# install customization libraries
library(viridis)
library(hrbrthemes)

payment_plot <- ggplot(payment, aes(fill = rental_methods,x = rental_methods, y = subscription)) + 
geom_col(position="stack", stat="identity") + 
labs(title = "Payment method",
     x = "Rental Method",
     y = "Number of rentals")

payment_plot

```
Unlike what could be assumed the single ride payment is the most used rental method. In proportion it is more than 88% ! Numerous reasons can be given for explaining such gap but maybe locals are not using as many as citybikes as tourists, or they are not aware that a monthly subscription will cost as much as the same as a single ride.

### How much sales ?
Assuming this data, it could be great to see how much tunrover is doing Velib on single ride rentals. Nevertheless it is not possible to correctly know how much it is for subscriptions as different subscription methods are existing at different fees. 

In the following part of the study we will assume that payment with credit card will only be one time payment that don't exceed 45 min. Time when you get charged extra fee.

A single ride is costing 3€ for a mechanical or an ebike. 

```{r}
turnover <- payment[1,] %>% #take only the first row 
            mutate(turnover = subscription * 3) 

#display the turnover for one day, create a new table with the date and the turnover
turnover_over_date <- dbGetQuery(conn, 'SELECT rental_methods, count(*) as subscription, count(*)*3 as                                          turnover, export_date
                                       FROM Velib_Data
                                       WHERE rental_methods == "CREDITCARD"')


          ```


## Local Analysis 

### What is the variation of bikes available within a day ?
In order to analyse which stations are lacking citybikes, first step consists in seeing how many bikes are available within a day to analyse peak hours. As each line refers to one transaction let's create a table with all the transaction done within one day. 

```{r}

```


Let's see now how these bikes are dispatched around the city. For such purpose we should display the list of bikes by descending order with number of bikes available. However this ranking will depend of peak hours. The more a station uner pressure the fewer bikes available. It consequently requires to timeframe days in 3 parts : morning (from 6:00 AM to 12:00AM), afternoon (from 12:01 AM to 6:00PM) and night (from 6:01 PM to 5:99 AM)


```{r}
# sorting the station per available bikes 


availability <- dbGetQuery(conn,'SELECT DISTINCT )

```
As you might know a lot of bikes are unavailable due to many reasons of dysfunctionment. However as you might have noticed the number of bikes available equals the number of mechanical and ebikes lastly reported in the station. But it does not mean that every bike is fully operational. 

```{r}

bike_plot

```




### Is subscription dominant over one time payment ? 
```{r}
# compute percentages over the mechanical bikes 


```